Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-22 manuscript/HIAs.qmd
crop terra and summary table

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
```

@@

naturetypes <- naturetypes |>
  # keep only wetlands and our selected nature types
  filter(hovedøkosystem == "våtmark",
         naturtype %in% c(advsNTs, alienNTS)) |>
  # calculate the areas (m2) of the polygons
  mutate(area = SHAPE |> st_area()) |>
  # the variable codes and values are all in the same column
  separate_rows(ninBeskrivelsesvariable, sep=",") |>
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) |>
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
  select(id = identifikasjon_lokalId,
         municipality = kommunenummer,
         year = kartleggingsår,
         mosaic = mosaikk,
         quality = lokalitetskvalitet,
         biodiversity = naturmangfold,
         condition = tilstand,
         natureType = naturtype,
         variable = NiN_variable_code,
         value = NiN_variable_value,
         area) |>
  st_transform(25832)

@@

```{r getNorwayOutline}
#| eval: true
outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
  st_transform(25832)
```

@@

# find the correct layer
#st_layers(path)
muni <- sf::read_sf(path, layer = "kommune") |>
  st_transform(25832)

@@

# claculate stats

# terrestrial area
terrestrial <- terrestrial |>
  mutate(area_t = geometry |> st_area())

@@

```{r makeMap}
# Make map to show where the three municipalities are
# First, get centroids
centroids <- muni3 |>
  st_centroid()

@@

tmap_save(methodsMap, "../figures/studyLocations.tiff",
          dpi= 1000,
          units = "cm",
          width = 18,
          height = 10)

saveRDS(methodsMap, "../figures/studyLocation.RDS")
```

```{r getMireData}
file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
# This is a big TIF, so I some lazy evalyation on the starts proxy 
tr2 <- tr |>
  st_transform(25833)

mire <- read_stars(file) |>
  st_downsample(100) |>
  st_as_stars()
plot(mire, col="yellow")

tr_mire  <- stars::read_stars(file, proxy=T) |>
  st_crop(tr) |>
  st_as_stars()
#Not working

# What is the EDM is actually utm 33
tr_mire  <- stars::read_stars(file, proxy=T)
st_crs(tr_mire) <- 25833
tr_mire  <- tr_mire |>
  st_crop(tr2) |>
  st_as_stars()

# Same result, different syntax:
#nf_mire  <- stars::read_stars(file, proxy=T)
#nf_mire <- nf_mire[nf]
#nf_mire <- st_as_stars(nf_mire)

# This is not working as it puts the cropped raster into the sea
tmap_mode("view")
tm_shape(st_downsample(tr_mire , 10))+
  tm_raster(alpha=.8)+
  tm_shape(outline)+
  tm_borders()

# trying terra
# convert municipal outline to vect via st
nf2 <- as(nf, "Spatial") |>
  terra::vect()

# crop and mask (very fast!)
nf_mire2 <- terra::rast(file) |>
  crop(nf2) |>
  terra::mask(nf2)

# Plot to check overlap
ggplot()+
  geom_spatraster(data = nf_mire2)+
  geom_spatvector(data = nf2,
                  fill = NA)
# The cropping and masking worked.

# I ilke the stars, sf and tmap combo better, so I return to stars
nf_mire3 <- nf_mire2 |>
  st_as_stars()

tmap_mode("plot")
tm_shape(st_downsample(nf_mire3 , 10))+
  tm_raster(alpha=.8
            )+
  tm_shape(outline)+
  tm_borders()

# Then I cut out the relevant part of the ecosystem map
nft <- nf |>
  terra::vect()

ras_crop <- terra::crop(ras_raw, shp)
ras_mask <- terra::mask(ras_crop, shp, touches = T)

nf_mire <- st_crop(EDM, st_bbox(nf))
tr_mire <- st_crop(EDM, tr)
me_mire <- st_crop(EDM, me)

  tm_shape(nf)+
  tm_borders(col="red", 
             lwd=3)+
tm_shape(st_downsample(nf_mire, 10))+
  tm_raster(col="myrmodell90pros.tif")

#tm_shape(st_downsample(EDM, 100))+
  #tm_raster(alpha=.3)+

@@

We chose three municipalities in Norway to test out the indicator @fig-studyLocation.
These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (TABLE)

@@

# calculate terrestrial area
terrestrial <- terrestrial |> 
  mutate(t_area_km = 
           round(units::drop_units(area_t * 1e-6))) |>
  as_tibble()

muni_tbl <- muni3 |>
  # calculate total area
  mutate(area_km = 
           round(
             units::drop_units(
               geom |> st_area() * 1e-6))) |>
  # make tibble for the coming join
  as_tibble() |>
  # paste inn terrestrial area
  left_join(terrestrial |> select(kommunenummer, t_area_km),
            keep=F) |>
  # area of survey
  left_join(dk2 |> select(Municipality, dk_area_km)) |>
  mutate(dk_percent = round((dk_area_km/t_area_km) * 100))

@@

| Municipality | Total terrestrial area (km2)                                                 | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area | Total mire area | Fraction mire area included in survey | Number of mire polygons in survey |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Nordre Follo | `r terrestrial |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` |                                                     |                 |                                       |                                   |
| Trondheim    | `r terrestrial |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                     |                 |                                       |                                   |
| Målselv      | `r terrestrial |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                     |                 |                                       |                                   |

>>>>

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
library(tidyterra)

myCRS <- 25832
```

@@

naturetypes <- naturetypes |>
  # keep only wetlands and our selected nature types
  filter(hovedøkosystem == "våtmark",
         naturtype %in% c(advsNTs, alienNTS)) |>
  # calculate the areas (m2) of the polygons
  mutate(area = SHAPE |> st_area()) |>
  # the variable codes and values are all in the same column
  separate_rows(ninBeskrivelsesvariable, sep=",") |>
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) |>
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
  select(id = identifikasjon_lokalId,
         municipality = kommunenummer,
         year = kartleggingsår,
         mosaic = mosaikk,
         quality = lokalitetskvalitet,
         biodiversity = naturmangfold,
         condition = tilstand,
         natureType = naturtype,
         variable = NiN_variable_code,
         value = NiN_variable_value,
         area) |>
  st_transform(myCRS) # Chosing this to match the EDM (se further down)

@@

```{r getNorwayOutline}
#| eval: true

outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
  st_transform(myCRS)
```

@@

# find the correct layer
#st_layers(path)
muni <- sf::read_sf(path, layer = "kommune") |>
  st_transform(myCRS)

@@

# calculate stats
# terrestrial area
terrestrial <- terrestrial |>
  mutate(area_t = geometry |> st_area())

@@

```{r makeMap}
#| eval: false

tmap_mode("plot")
# Make map to show where the three municipalities are
# First, get centroids
centroids <- muni3 |>
  st_centroid()

@@

# tmap_save(methodsMap, "../figures/studyLocations.tiff",
#           dpi= 1000,
#           units = "cm",
#           width = 18,
#           height = 10)
# 
# saveRDS(methodsMap, "../figures/studyLocation.RDS")
```

```{r mireDataTestCrop}
file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
# This is a big TIF!

mire_test <- read_stars(file, proxy = T) |>
  setNames("mire") |>
  st_downsample(100) |>
  mutate(mire = na_if(mire, 0)) |>
  st_as_stars()

mire_test_nf <-  mire_test |>
  st_crop(nf)

# The cropping is correct
tmap_mode("plot")
tm_shape(mire_test)+
  tm_raster(palette = "red")+
  tm_shape(outline)+
  tm_borders()+
  tm_shape(mire_test_nf)+
  tm_raster(col="red")
```

```{r mireNF-error}
#| eval: false
# Then we do the same, but without downsampling

mire_nf <- read_stars(file, proxy = T) |>
  setNames("mire") |>
  #mutate(mire = na_if(mire, 0)) |>
  st_crop(nf) |>
  st_as_stars()

# The cropped raster is for some reason far into the ocean. 
# The cells have values, so it seems like there is something 
# that happens after the cropping
tmap_mode("plot")
tm_shape(mire_test)+
  tm_raster(palette = "red")+
  tm_shape(outline)+
  tm_borders()+
  tm_shape(mire_nf)+
  tm_raster(col="red")

st_crs(mire_nf) <- 25832
```

```{r mireTerra}
# trying terra
# convert municipal outline to vect via st
nf_vect <- as(nf, "Spatial") |>
  terra::vect()

# crop and mask (very fast!)
mire_nf_terra <- terra::rast(file) |>
  crop(nf_vect) |>
  terra::mask(nf_vect)

# Plot to check overlap
ggplot()+
  geom_spatraster(data = mire_nf_terra)+
  geom_spatvector(data = nf_vect,
                  fill = NA)
# The cropping and masking worked. 
#tmap_mode("plot")
#tm_shape(mire_test)+
#  tm_raster(palette = "red")+
#  tm_shape(outline)+
#  tm_borders()+
#  tm_shape(mire_nf_terra)+
#  tm_raster(col="red")

# I ilke the stars, sf and tmap combo better, so I return to stars
mire_nf_stars <- mire_nf_terra |>
  st_as_stars()

tm_shape(mire_test)+
  tm_raster(palette = "red")+
  tm_shape(outline)+
  tm_borders()+
  tm_shape(mire_nf_stars)+
  tm_raster(col="red")
# OK

ggplot()+
  geom_stars(data = mire_nf_stars)+
  coord_equal()+
  theme_void()+
  scale_fill_viridis_c()
```

```{r mireArea}
mireArea <- mire_nf_terra |>
  global(c("mean", "sum"), na.rm=T) |>
  add_column("Municipality" = "Nordre Follo") |>
  mutate(mirePercent = round(mean*100, 1),
         sum = sum/1e+4) |>
  dplyr::rename(mire_km2 = sum)

mire_nf_terra_dk <- mire_nf_terra |>
  terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
  global("sum", na.rm=T) |>
  mutate(mireInSurvey_km2 = sum/1e+4) |>
  add_column(Municipality = "Nordre Follo")

@@

We chose three municipalities in Norway to test out the indicator @fig-studyLocation.
These municipalities are spread over xxx km north to south, and also differ in the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-variables).

```{r}
# calculate terrestrial area
terrestrial <- terrestrial |> 
  mutate(t_area_km = 
           round(units::drop_units(area_t * 1e-6))) |>
  as_tibble()

# number and area of mire polygons
nature3_tbl <- nature3 |>
  group_by(Municipality) |>
  summarise(n = n()) |>
  as_tibble()

muni_tbl <- muni3 |>
  # calculate total area
  mutate(area_km = 
           round(
             units::drop_units(
               geom |> st_area() * 1e-6))) |>
  # make tibble for the coming join
  as_tibble() |>
  # paste inn terrestrial area
  left_join(terrestrial |> select(kommunenummer, t_area_km),
            keep=F) |>
  # area of survey
  left_join(dk2 |> select(Municipality, dk_area_km)) |>
  mutate(dk_percent = round((dk_area_km/t_area_km) * 100)) |>
  # number of polygons
  left_join(nature3_tbl |> select(Municipality, n)) |>
  # total mire area and %
  left_join(mireArea |> select(Municipality, mirePercent, mire_km2)) |>
  # % mire inside survey coverage map
  left_join(mire_nf_terra_dk |> select(Municipality, mireInSurvey_km2)) |>
  mutate(mireInSurvery_percent = round(mireInSurvey_km2/mire_km2*100, 2))

saveRDS(muni_tbl, "../data/municipality_stats.RDS")

@@

muni_tbl <- readRDS("../data/municipality_stats.RDS")

@@

| Municipality | Total terrestrial area (km^2^)                                            | \% of terrestrial area surveyed                                            | \% open mires in relation to total terrestrial area                         | Total mire area (km^2^)                                                  | %of mire area inside survey area                                                      | Number of mire polygons in survey                                 |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Nordre Follo | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(t_area_km)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(dk_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mirePercent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mire_km2)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(mireInSurvery_percent)` | `r muni_tbl |> filter(Municipality == "Nordre Follo") |> pull(n)` |
| Trondheim    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(t_area_km)`    | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(dk_percent)`    |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Trondheim") |> pull(n)`    |
| Målselv      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(t_area_km)`      | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(dk_percent)`      |                                                                             |                                                                          |                                                                                       | `r muni_tbl |> filter(Municipality == "Målselv") |> pull(n)`      |
